How Optimal is Word-Referent Identification Under Multimodal Uncertainty?

### 2019-03-19 Writeup/ms.tex
revisions

\subsection{Descriptive model and analysis of
sub-optimality}\label{descriptive-model-and-analysis-of-sub-optimality}

@@

\paragraph{Normative models}\label{normative-models}

Figure~\ref{fig:bimodal} compares the predictions of the normative
models against human responses. Remember that the normative models use
variance estimates from the unimodal conditions (where people see input
from only one modality) to predict data in the bimodal condition (where
people see input from both modalities). Figure \ref{fig:bimodal} shows
the models' predictions comapred to human data.

\paragraph{Descriptive model}\label{descriptive-model}

In the descriptive model, all parameters are fit to human responses in
the bimodal condition. The prediction of this model are shown in Figure
\ref{fig:bimodal}. Note that the descriptive model explained almost all
the variance, which makes it a reasonble proxy for human real
performance in the bimodal condition. Thus, to determine the extent to
which human data deviated from optimality, we comapred the descriptive
and optimal models follwoing the procedure outlined above (section XX).

@@

\subsubsection{Analysis of optimality}\label{analysis-of-optimality}

We analyzed if the cue combination was performed in an optimal way, or
if there was a systematic preference for one modality when making
decisions in the bimodal condition. As explained above
(\ref{fig:subOptim}), modality preference can be characterized formally
as a deviation from the decision threshold predicted by the optimal
model. The results in Figure \ref{fig:bias} (top) show both the decision
threshold derived from the descriptive model (in black) and the decision
threshold predicted by the optimal model (in red). The deviation from
optimality is compared to two hypothetical cases of modality preference
(dotted lines). We found that the descriptive and optimal decision
thresholds were almost identical. Indeed, non-parametric resampling of
the data showed no evidence of a deviation from the optimal prediction
(Figure~\ref{fig:bias}, bottom).

@@

We found that the optimal model explained largely more variance than
what can be explained with the auditory or the visual models alone. in
addition, the high correlation with the mean human data suggests that
the population was, overall, near-optimal. However, we see in Figure
\ref{fig:bimodal} that the responses deviated systematically from the
optimal prediction in that they were slightly pulled toward chance
(i.e., the probability 0.5). This fact is due to the increase in the
value of the variance associated with each modality (as can be noted
from Table \ref{tab:param}). Note however that, despite this increase in
randomness, our analysis of modality preference showed that the relative
values of these variances were not different (Figure~\ref{fig:bias}),
meaning that there was no evidence for a modality preference. Thus, 1)
There was a simultaneous increase in the values of the auditory and
visual variances in the bimodal condition compared to the unimodal
condition, meaning that the bimodal input lead to an increase in
response randomness, and 2) this increased randomness did not affect the
relative weighting of both modalities, i.e., the participants was
weighting modalities according to the relative reliability predicted by
the optimal model. This situation corresponds to the first case of
sub-optimality described in Figure~\ref{fig:subOptim}.

@@

We fit a model for each modality. Figure \ref{fig:unimodal} shows human
responses together with their best fits. The visual data is a
replication of the visual data in Experiment 1. As for the auditory
data, in contrast to Experiment 1, responses were flatter, showing more
uncertainty.

@@

We used the values derived from the unimodal condition (Table
\ref{tab:param}) to construct the visual, auditory and optimal models.
In addition, we fit a descriptive model which allows us to assess real
human performance in this condition. Figure~\ref{fig:bimodal} compares
the predictions of these models to human responses. Note that, in
contrast to Experiment 1, the visual model explained more variance than
the auditory model did.

\subsubsection{Analysis of optimality}\label{analysis-of-optimality-1}

Figure~\ref{fig:bias} (top) shows that the participants' decision
threshold deviated from optimality, and that this deviation was biased
towards the visual modality (the non-noisy modality). Indeed
non-parametric resampling of the data showed a decrease in the value of
the slope in the descriptive model compared to the optimal model
(Figure~\ref{fig:bias}, bottom).

@@

We found, similar to Experiment 1, that the optimal model explained more
variance than the auditory or the visual models alone (Discuss the high
correlation, and why this not in itself a necessary but not sufficient
indicator of optimality). We also found a similar discrepancy between
the descriptive and optimal models as precision dropped for both the
auditory and the visual modalities. As for the weighting scheme used by
participants, contrary to Experiment 1 where modalities were weighted
according to their relative reliability, we found in this experiment
that the visual modality had a greater weight than what was expected
from its relative reliability. This situation corresponds to the second
case of sub-optimality described in Figure \ref{fig:subOptim}.

In Experiment 2, we tested the case of added background noise to the
auditory modality. In Experiment 3, we test the case of added noise to
the visual modality.

@@

In this Experiment, we added background noise to the visual modality.
Similar to Experiment 2, we were interested to know if participants
would treat this new source of uncertainty as predicted by the optimal
model, that is, according to the following weighting scheme:
\[\beta_a \propto \frac{1}{\sigma^2_{A}}\]
\[\beta_v \propto \frac{1}{\sigma^2_{V}+\sigma^2_{N_V}}.\] The
alternative hypothesis is that noise in the visual modality would lead
to a preference for the auditory input, just like noise in the auditory
modality lead to a preference for the visual input in Experiment 2.

@@

\subsection{Results and analysis}\label{results-and-analysis-1}

@@

For the auditory modality, our parameter estimates were. For the visual
modality. Figure~\ref{fig:unimodal} shows responses in the unimodal
conditions as well as the corresponding fits. The auditory data is a
replication of the auditory data in Experiment 1. As for the visual
data, we found that, in contrast to Experiment 1 and 2, responses were
flatter, showing much more uncertainty.

@@

\paragraph{Normative models}\label{normative-models-1}

Figure~\ref{fig:bimodal} compares the predictions of the visual,
auditory and optimal models to human responses. \#\#\#\# Descriptive
model Cross-validation using half the responses to predict the other
half yielded \(r^2=\)

\paragraph{Modality preferences}\label{modality-preferences}

Participants' decision threshold suggested a preference for the auditory
modality (the non-noisy modality). Indeed non-parametric resampling of
the data showed an increase in the value of the slope in the descriptive
model compared to the optimal model (Figure~\ref{fig:bias}).

@@

We found that the optimal model accounted for almost all the variance.
However, whereas in previous experiments the optimal model explained
more variance than the auditory or the visual models, here the auditory
model explained at least as much variance. Thus, though participants
were still sensitive to variation in the noisy visual data in the
unimodal condition, they tended to ignore this information in the
bimodal condition, and relied almost exclusively on the non-noisy
auditory modality. The reason why we saw this (floor) effect when we
added noise to the visual modality (Experiment 3), and not when we added
noise to the auditory modality (Experiment 2), is the fact that our
visual stimuli were originally perceived less categorically and with
less certainty than the auditory stimuli. This fact made it more likely
for the visual categorization function to become flat and uninformative
after a few drops in precision due to noise on the one had, and to the
additional randomness induced by the bimodal presentation on the other
hand.

The general finding corresponds to the third case of sub-optimality
described in Figure~\ref{fig:subOptim}. Indeed, precision dropped for
both modalities in the bimodal condition compared to the unimodal
condition. But the drop was much greater for the visual modality,
resulting in a much lower weight assigned to it than what is expected
from its reliability. Therefore, just like participants over-relied on
the visual modality when the auditory modality was noisy (Experiment 2),
they also over-relied on the auditory modality when the visual modality
was noisy (Experiment 3).

The past three experiments studied the problem of cue combination at the
population level --- The models were fit to the data aggregted across
all participants. However, optimality at the population level could be
misleading. Imagine, for instance, that in the bimodal condition half
the participants relied exclusively on the visual modality, whereas the
other half relied exclusively on the auditory modality. This case could
still lead to an aggregate behavior which appears optimal, but this
optimality would be spurious. Thus, we need to test cue integration at
the individual level as well. As we noted earlier, in previous
experiments we did not have enough statistical power to fit a different
model for each participant. In Experiment 4, below, we re-ran Experiment
1 (where we found near optimal behavior at the population level) while
extending its length, allowing us to collect the number of datapoints
necessary to fit the models for each participant.

@@

We recruited a planned sample of N=XX participants from Amazon
Mechanical Turk. Only participants with US IP addresses and a task
approval rate above 99\% were allowed to participate. They were paid at
an hourly rate of \$6/hour. Participants were excluded if they reported
having experienced a technical problem of any sort during the online
experiment (N=XX), or if they had less than 75\% accurate responses on
the unambiguous training trials (N=XX). The final sample consisted of N
= XX participants. {[}I should also add how many participants I excluded
because their models did not converge{]}. All participants provided
informed consent before taking the
experiment.\footnote{The sample size and exclusion criteria were specified in the pre-registration at https://osf.io/h7mzp/.}

@@

The design and procedure were similar to Experiment 1. The only
difference was that we increased the number of datapoints elicited per
subject from 70 to 300.

We fit both population-level models to the aggregated data (as in
Experiment 1), and individual models to data from each participant. For
the populaton-level models, the results are shown in Figure
\ref{fig:unimodal}, Figure \ref{fig:bimodal}, and Table \ref{tab:param}.
We note that the results mirrors the patterns obtained in Experiment 1.

At the individual level, we computed the distribution of optimal
weighing relative to the fitted (i.e., descriptive) weighing and we show
this distribution in Figure \ref{fig:weights}. The distribution centers
around the optimal prediction (i.e., 1) with a median value of XX
{[}XX,XX{]}. The variance of the distribution indicates that, while most
participants were near optimal in their cue combination, part of the
participant tended to rely predominantly on the auditory modality and
another part tended to rely predominantly on the visual modality.

@@

This experiment was an extension to Experiment 1. Through collecting
larger-size data per subject, we were able to analyze the cue
combination optimlity, not only at the population level, but also at the
individual level. The population-level analysis replicated the results
of Experiment 1. The individual-level analysis showed that participants'
performance approximated a normal distribution centered around the
optimal combination. The large variance of this distribution indicate
that some participants tend to over-rely on the auditory modality and
others tend to over-rely on the visual modality.

@@

In the current paper, we explored word identification under uncertainty
about both words and their referents. We conducted an ideal observer
analysis of this task whereby a model provided predictions about how
information from each modality should be combined in an optimal fashion.
The predictions of the model were tested in a series of three
experiments where instances of both the form and the meaning were
ambiguous with respect to their category membership only (Experiment 1),
when instances of the form were perturbed with additional background
noise (Experiment 2), and when instances of the referent were perturbed
with additional visual noise (Experiment 3). We discuss the findings of
these studies first with respect to our ideal observer model and
inferences about optimality and second with respect to their
implications for word identification more generally.

@@

In all of our experiments, at the highest level, we found generally
optimal behavior. Quantitatively speaking, the optimal model accounted,
respectively, for xx, xx, and xx of the variance in mean responses. When
compared to the predictions of the visual or the auditory models,
participants generally relied on both modalities to make their decisions
in the bimodal condition. Indeed, in Experiment 1 and 2, the optimal
model accounted for more variance in mean responses than the auditory or
the visual models did. In Experiment 3, participants appeared to rely on
one modality, but this was likely a floor effect, due to the fact that
noise made the visual input barely perceptible. Further, in Experiment
1, which did not involve background noise, participants not only relied
on both modalities, but generally weighted these modalities according to
the predictions of the optimal model, that is, according to their
relative reliability. At the individual level, however, we found
evidence of a between-subject variation: Some participants relied more
on the visual modality, whereas others relied more on the auditory
modality.

>>>>

\subsection{Descriptive model and analysis of
(sub-)optimality}\label{descriptive-model-and-analysis-of-sub-optimality}

@@

Figure \ref{fig:bimodal} compares the predictions of the normative and
descriptive models against human responses. Remember that the normative
models use the parameters estimated from the unimodal conditions (where
people see input from only one modality) to predict behavior in the
bimodal condition (where people see input from both modalities). The
descriptive model has a similar structure than the optimal model, but is
directly fit to human responses in the bimodal condition in order to
allow us to assess deviation from optimaity.

We found, through comparing the correlation values, that the optimal
model explained more variance than the visual and auditory models did.
However, the optimal model was not perfect: It explained less variance
than the descriptive model did, which indicates a deviation from
optimality. To investigate this deviation, we compare the parameter
values of the optimal model to the values obtaied in the decriptive
model (Table
\ref{tab:param}).\footnote{Note that the descriptive model explained almost all the variance in mean responses, which makes it a reasonble proxy for human real performance in the bimodal condition.}.
We note an increase in \emph{both} the auditory and visual variances.
This increase in noise is compatible with the fact that human responses
appear to be pulled towards chance (i.e., the value O.5) when compared
to the optimal model (see \ref{fig:bimodal}). Below we investigate if
this deviation from optimality can be related to the cue combination
strategy.

\subsubsection{Cue combination}\label{cue-combination}

We analyzed if the cue combination was performed in an optimal way, or
if there was a systematic preference for one modality when making
decisions in the bimodal condition. As explained in \ref{fig:subOptim},
modality preference can be characterized formally as a deviation from
the decision threshold predicted by the optimal model. The results in
Figure \ref{fig:bias} (top) show both the decision threshold derived
from the descriptive model (in black) and the decision threshold
predicted by the optimal model (in red). We found that the descriptive
and optimal decision thresholds were almost identical. Indeed,
non-parametric resampling of the data showed no evidence of a deviation
from the optimal prediction (Figure~\ref{fig:bias}, bottom).

@@

@@

This experiment studied the way participants combine multimodal
information to recognize novel words. We found that the optimal model
explained more variance than the auditory or the visual models did,
indicating that particiapants take into account both the auditory and
visual cues when making a decision. That said, Figure \ref{fig:bimodal}
shows that the participants deviated slightly --- but systematically---
from the optimal prediction in that they were slightly pulled toward
chance (i.e., the probability 0.5). This fact was captured by the
increase in the value of the variance associated with each modality (as
can be noted from Table \ref{tab:param}). Note, however, that despite
this increase response randomness in the bimodal condition, our analysis
of modality preference showed that the \emph{relative} values of these
variances were not different (Figure~\ref{fig:bias}), meaning that there
was no evidence for a modality preference.

To sum up, 1) the participants used both the auditory and visual
information, 2) they responded slightly more randomly that what was
predicted, but 3) this increased randomness was general and did not
influence the cue combination strategy, i.e., the participants still
weighted modalities according to their relative reliability as predicted
by the optimal model. This situation corresponds to the first case of
sub-optimality described in Figure~\ref{fig:subOptim}.

@@

We fit a model for each modality. Figure \ref{fig:unimodal} shows human
responses together with their best fits. The visual data is a
replication of the visual data in Experiment 1. The auditory data, in
contrast, were flatter, showing more uncertainty.

@@

We used the values derived from the unimodal condition to construct the
visual, auditory and optimal models. In addition, we fit a descriptive
model which allowed us to assess real human performance in this
condition. Figure \ref{fig:bimodal} shows that, similar to Experiment 1,
the optimal model explained more variance than the auditory and visual
models did (note, however, that the visual model explained more variance
than the auditory model did). Also similar to Exeperiment 1, the values
of the variances increased in the bimodal condition (Table
\ref{tab:param}).

\subsubsection{Cue combination}\label{cue-combination-1}

Here we investigated whethere the observed increase in the auditory and
visual variances affected the relative weighting of the corresponding
modalities. Figure~\ref{fig:bias} (top) shows that the participants'
decision threshold deviated from optimality, and that this deviation was
biased towards the visual modality (the non-noisy modality). Indeed
non-parametric resampling of the data showed a decrease in the value of
the slope in the descriptive model compared to the optimal model
(Figure~\ref{fig:bias}, bottom).

@@

Experiment 2 tested audi-visual combination in the case where the
auditory input was noisy. We found, similar to Experiment 1, that the
optimal model explained more variance than the auditory or the visual
models did. In other words, despite additional noise, participants still
used information from the noisy modality to recognize words. We also
found a similar discrepancy between the descriptive and optimal models
as response randomness increased along both the auditory and the visual
modalities. As for the relative weighting, and contrary to Experiment 1
where modalities were weighted optimally, we found in this experiment
that the visual modality had a greater weight than what was expected
from its relative reliability. This situation corresponds to the second
case of sub-optimality described in Figure \ref{fig:subOptim}.

Whereas in Experiment 2 we tested the case of added background noise to
the auditory modality, in Experiment 3 we test the case of added noise
to the visual modality.

@@

Similar to Experiment 2, we were interested to know if participants
would treat additional uncertainty as predicted by the optimal model,
that is, according to the following weighting scheme:
\[\beta_a \propto \frac{1}{\sigma^2_{A}}\]
\[\beta_v \propto \frac{1}{\sigma^2_{V}+\sigma^2_{N_V}}.\] The
alternative hypothesis is that noise in the visual modality would lead
to a preference for the auditory input, just like noise in the auditory
modality lead to a preference for the visual input in Experiment 2.

@@

@@

Figure \ref{fig:unimodal} shows responses in the unimodal conditions as
well as the corresponding fits. The auditory data is a replication of
the auditory data in Experiment 1. As for the visual data, we found
that, in contrast to Experiment 1 and 2, responses were flatter, showing
much more uncertainty.

@@

Figure \ref{fig:bimodal} shows that almost all the variance was captured
by the auditory model alone, the addition of visual information in the
optomal model did not improve the prediction of human responses. Similar
to Exeperiment 1 and 2, the values of the variances increased in the
bimodal condition (Table \ref{tab:param}).

\subsubsection{Cue Combination}\label{cue-combination-2}

Figure \ref{fig:bias} indicates that the decision threshold was biased
towards the auditory modality (the non-noisy modality). Indeed
non-parametric resampling of the data showed an increase in the value of
the slope in the descriptive model compared to the optimal model
(Figure~\ref{fig:bias}).

@@

Experiment 3 tested audi-visual combination in the case where the visual
input was noisy. Whereas in previous experiments the optimal model
explained more variance than the auditory or the visual models did, here
the auditory model alone explained almost all the variance. In other
words, though participants were sensitive to variation in the noisy
visual input when presented in isolation (as shown in Figure
\ref{fig:unimodal}), they tended to ignore this information when the
visual input was presented simultaneusel with the auditory input (i.e.,
in the bimodal condition). Instead, they relied almost exclusively on
the non-noisy auditory
modality.\footnote{The reason why we saw this (floor) effect when we added noise to the visual modality (Experiment 3), and not when we added noise to the auditory modality (Experiment 2), is the fact that our visual stimuli were originally perceived less categorically and with less certainty than the auditory stimuli (see Experiment 1 in \\ref{fig:unimodal}). This fact made it more likely for the visual categorization function to become flat and uninformative after a few drops in precision due to noise on the one had, and to the additional randomness induced by the bimodal presentation on the other hand.}

This finding corresponds to the third case of sub-optimality described
in Figure~\ref{fig:subOptim}. Indeed, precision dropped for both
modalities in the bimodal condition compared to the unimodal condition.
But the drop was much greater for the visual modality, resulting in a
much lower weight assigned to it than what is expected from the optimal
model. Therefore, just like participants over-relied on the visual
modality when the auditory modality was noisy (Experiment 2), they also
over-relied on the auditory modality when the visual modality was noisy
(Experiment 3).

So far we have studied the problem of cue combination at the population
level --- the models were fit to the data aggregated across all
participants. However, it is important to investigate individual
variability, especially in Experiment 1 where we reported optimal cue
combination. In fact, it is possible that a large part of the
participants relied primarily on the visual modality and another part on
the auditory modality. Such an extreme individual variability could
possibly lead to an aggregate beahvior which appears optimal, but such
optimality would be spurious. In order to rule out this extreme case, we
need to examine the distribution of the specific cue combibination
strategies followed by the participants.

As we noted earlier, we did not have enough statistical power to fit a
different model for each participant. Thus, in Experiment 4, below, we
reran Experiment 1 while extending its length, allowing us to collect
the number of datapoints necessary to fit the models for each
participant.

@@

A planned sample of \(N =\) 100 participants was recruited online
through Amazon Mechanical Turk. We used the same exclusion criteria as
in both previous experiments. \(N =\) 2 participants were excluded
because they reported having a technical problem, and N=10 participants
were excluded because they had less than 50\% accurate responses on the
unambiguous training trials. The final sample consisted of \(N =\) 88
participants.

We recruited a planned sample of \(N=\) 50 participants from Amazon
Mechanical Turk. Only participants with US IP addresses and a task
approval rate above 99\% were allowed to participate. They were paid at
an hourly rate of \$6/hour. Participants were excluded if they reported
having experienced a technical problem of any sort during the online
experiment (\(N=\) 0 ), or if they had less than 75\% accurate responses
on the unambiguous training trials (\(N=\) 7). The final sample
consisted of \(N =\) 43 participants. All participants provided informed
consent before taking the
experiment.\footnote{The sample size, exclusion criteria and the main analysese were pre-registered at https://osf.io/h7mzp/.}

@@

The design and procedure were similar to Experiment 1. The only
difference was that we increased the number of responses elicited per
subject from 70 to 300.

\subsection{Results}\label{results-2}

\subsubsection{Unimodal and Bimodal
conditions}\label{unimodal-and-bimodal-conditions}

In order to replicate the analysis of Experiment 1, we started by
fitting population-level models to the aggregated data. Indeed, we found
that the results --- as shown in Figure \ref{fig:unimodal}, Figure
\ref{fig:bimodal}, and Table \ref{tab:param} --- mirror closely the
patterns obtained in Experiment 1.

\subsubsection{Cue combinaton}\label{cue-combinaton}

We analyzed cue combination at the individual level. For each
participant, we computed the optimal weighing,
\(\frac{\sigma^2_{A}}{\sigma^2_{V}}\), relative to the fitted (i.e.,
descriptive) weighing, \(\frac{\sigma^2_{Ab}}{\sigma^2_{Vb}}\). We show
the resulting distribution in Figure \ref{fig:weights}. We found that
the distribution centered around the optimal prediction (i.e., 1) with a
median value of 1.07 {[}\texttt{weight.median\_low},
\texttt{weight.median\_up}{]}.

@@

This experiment was an extension to Experiment 1. Through collecting
larger-size data per subject, we were able to analyze the cue
combination optim\&lity, not only at the population level, but also at
the individual level. The population-level analysis replicated the
results of Experiment 1. The individual-level analysis showed that the
distribution of cue combination scores had a unimodal shape
(approximating a normal distribution) centered around the optimal
combination. This finding rules out the hypotheis that optimality at the
population level (obtained in Experiment 1) is a spurious finding, i.e.,
only obtained via the aggregation of sub-otimal strategies. That said,
the large variance of this distribution indicates that some participants
tend to over-rely on the auditory modality and others tend to over-rely
on the visual modality (beyond measurement errors??).

@@

In the current paper, we explored word identification under uncertainty
about both words and their referents. We conducted an ideal observer
analysis of this task whereby a model provided predictions about how
information from each modality should be combined in an optimal fashion.
The predictions of the model were tested in a series of four experiments
where instances of both the form and the meaning were ambiguous with
respect to their category membership only (Experiment 1 and 4), when
instances of the form were perturbed with additional background noise
(Experiment 2), and when instances of the referent were perturbed with
additional visual noise (Experiment 3). We discuss the findings of these
studies first with respect to our ideal observer model and inferences
about optimality and second with respect to their implications for word
identification more generally.

@@

In all of our experiments, and when compared to the predictions of the
visual or the auditory models, participants generally relied on both
modalities to make their decisions in the bimodal condition. Indeed, in
Experiment 1 and 2, the optimal model accounted for more variance in
mean responses than the auditory or the visual models did. In Experiment
3, participants appeared to rely on one modality, but this was likely a
floor effect, due to the fact that noise made the visual input barely
perceptible. Further, in Experiment 1 and 4, which did not involve
background noise, participants not only relied on both modalities, but
generally weighted these modalities according to the predictions of the
optimal model, that is, according to their relative reliability. At the
individual level, however, we found evidence of a between-subject
variation: Some participants relied more on the visual modality, whereas
others relied more on the auditory modality.
